---
title: "Rare UTR Variants — WGS Annotation Pipeline"
author: "rareUTRvariants Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  eval    = FALSE   # Set eval=TRUE for each step when running sequentially
)
```

# Overview

This pipeline processes whole-genome sequencing (WGS) data to identify and annotate
rare UTR variants using gnomAD v3.1.2 population frequency data.

The main goal is to start from cohort-level variant calls, convert them into a
standardized hg38-based representation, and then attach population frequency fields
from gnomAD for downstream filtering. A key design principle of the workflow is that
all matching steps are based on a normalized variant identifier (`canonical_id`), which
helps reduce ambiguity when the same variant can be represented in more than one way.

The pipeline starts from a pre-merged input table (`input_variants.xlsx`) containing
variants from one or more WGS cohorts, with the following sheets:

- `Analysis Info` — run metadata
- `Quality Data` — per-variant quality metrics
- `Affected only all` — variant calls for affected individuals (primary input)

**Workflow summary:**

| Step | Tool | Description |
|---|---|---|
| 1 | R | Generate BED file from hg19 coordinates |
| 2 | Bash | LiftOver: hg19 → hg38 |
| 3 | R | Integrate hg38 coordinates into main table |
| 4 | R | Build VCF with anchor bases for indels |
| 5 | Bash | Normalize VCF (bcftools norm) |
| 6 | R | Build canonical_id bridge table |
| 7 | R | Add canonical_id to main table |
| 8 | R | Extract unique variants |
| 9 | R | Generate per-chromosome region files |
| 10 | Bash | Query gnomAD VCF files (bcftools query) |
| 11 | R | Combine gnomAD hit files |
| 12 | R | Annotate main table with gnomAD frequencies |
| 13 | R | Add variant_num (patient recurrence) → ready_for_filtration |

**Required input files:**

| File | Description |
|---|---|
| `input_variants.xlsx` | Pre-merged WGS variant table (`Affected only all` sheet) |
| `hg19ToHg38.over.chain` | UCSC LiftOver chain file |
| `reference.fa` + `reference.fa.fai` | GRCh38 reference genome (FASTA + index) |
| `gnomad.genomes.v3.1.2.sites.chrN.vcf.bgz` | gnomAD v3.1.2 genome VCF files (per chromosome) |

**Output files:**

| File | Content |
|---|---|
| `step01_bed_hg19.txt` | BED file with hg19 coordinates |
| `step02_hg38_coords.xlsx` | Main table with hg38 coordinates |
| `step03_raw.vcf` | Raw VCF (hg38) |
| `step04_normalized.vcf.gz` | Normalized and indexed VCF |
| `step05_bridge.csv` | canonical_id bridge table |
| `step06_with_canonical_id.xlsx` | Main table with canonical_id |
| `step07_unique_variants.csv` | Unique variant list |
| `step08_gnomad_annotations.csv` | gnomAD frequency annotations |
| `step09_annotated.xlsx` | Final annotated table with `ready_for_filtration` sheet |

---

# Required Packages

```{r packages}
# R packages
library(readxl)
library(dplyr)
library(openxlsx)
library(Biostrings)
library(BSgenome)
library(stringr)
library(tidyr)
library(purrr)
```

---

# Step 1 — Generate BED File (R)

hg19 coordinates from the input table are converted to BED format for use with UCSC liftOver.
BED format is 0-based, so `Start - 1` is applied to the start position. The `name`
column stores a compact variant label (`Chr_Start_Ref_Alt`) that will later be used to
link the liftOver output back to the original table.

```{r step01-bed}
df <- read_excel("input_variants.xlsx", sheet = "Affected only all")

bed <- df %>%
  transmute(
    chrom = paste0("chr", Chr),
    start = as.integer(Start) - 1L,
    end   = as.integer(Start),
    name  = paste(Chr, Start, Ref, Alt, sep = "_")
  ) %>%
  distinct()

dir.create("liftover_files", showWarnings = FALSE)
write.table(bed, "liftover_files/step01_bed_hg19.txt",
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

cat("BED rows written:", nrow(bed), "\n")
```

---

# Step 2 — LiftOver hg19 → hg38 (Bash)

Run UCSC liftOver to convert coordinates from hg19 to hg38.
Download liftOver binary and chain file from [UCSC](https://hgdownload.soe.ucsc.edu/admin/exe/).

```{bash step02-liftover, engine="bash"}
liftOver \
  liftover_files/step01_bed_hg19.txt \
  hg19ToHg38.over.chain \
  liftover_files/step02_mapped_hg38.txt \
  liftover_files/step02_unmapped.txt

echo "Mapped:"
wc -l liftover_files/step02_mapped_hg38.txt

echo "Unmapped:"
wc -l liftover_files/step02_unmapped.txt
```

---

# Step 3 — Integrate hg38 Coordinates (R)

The hg38 coordinates from liftOver are merged back into the main variant table.
A `canonical_key` (Chr_Start_Ref_Alt) is used as the join key. At this stage, the key is
only used to reconnect the mapped coordinates to the original records; it is not yet the
final normalized variant identifier used for gnomAD matching.

```{r step03-integrate}
df <- read_excel("input_variants.xlsx", sheet = "Affected only all")

hg38 <- read.table("liftover_files/step02_mapped_hg38.txt",
                   sep = "\t", header = FALSE,
                   col.names = c("chrom38", "start38", "end38", "name")) %>%
  mutate(
    Chr38   = sub("chr", "", chrom38),
    Start38 = as.integer(end38)
  ) %>%
  select(name, Chr38, Start38)

df <- df %>%
  mutate(canonical_key = paste(Chr, Start, Ref, Alt, sep = "_")) %>%
  left_join(hg38, by = c("canonical_key" = "name"))

wb <- createWorkbook()
addWorksheet(wb, "variants_hg38")
writeData(wb, "variants_hg38", df)
saveWorkbook(wb, "step02_hg38_coords.xlsx", overwrite = TRUE)

cat("Rows with hg38 coords:", sum(!is.na(df$Start38)), "\n")
cat("Rows missing hg38 coords:", sum(is.na(df$Start38)), "\n")
```

---

# Step 4 — Build VCF with Anchor Bases (R)

For indels, a 1-bp anchor base is prepended to REF and ALT alleles as required by VCF specification.
The `Biostrings` package is used to fetch the anchor base from the reference genome. This
step is essential because indels must be represented in a valid VCF-compatible form before
normalization and external database matching; otherwise, the same biological variant may
fail to match the corresponding gnomAD record.

```{r step04-vcf}
library(Biostrings)

df <- read_excel("step02_hg38_coords.xlsx", sheet = "variants_hg38")

# Load reference genome
ref_genome <- readDNAStringSet("reference/reference.fa")
names(ref_genome) <- sub(" .*", "", names(ref_genome))

get_anchor_base <- function(chrom, pos, ref_genome) {
  chr_name <- paste0("chr", chrom)
  if (!chr_name %in% names(ref_genome)) return(NA_character_)
  as.character(subseq(ref_genome[[chr_name]], start = pos, end = pos))
}

df_vcf <- df %>%
  filter(!is.na(Chr38), !is.na(Start38)) %>%
  mutate(
    is_indel    = nchar(Ref) != nchar(Alt),
    anchor_base = ifelse(is_indel,
                    mapply(get_anchor_base, Chr38, Start38 - 1, MoreArgs = list(ref_genome)),
                    NA_character_),
    VCF_POS     = ifelse(is_indel, Start38 - 1L, Start38),
    VCF_REF     = ifelse(is_indel, paste0(anchor_base, Ref), Ref),
    VCF_ALT     = ifelse(is_indel, paste0(anchor_base, Alt), Alt)
  )

# Save pre-bridge table
write.csv(df_vcf, "step03_pre_bridge.csv", row.names = FALSE)

# Write raw VCF
vcf_lines <- c(
  "##fileformat=VCFv4.2",
  paste0("##reference=GRCh38"),
  "#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO"
)
vcf_data <- df_vcf %>%
  transmute(
    CHROM  = paste0("chr", Chr38),
    POS    = VCF_POS,
    ID     = ".",
    REF    = VCF_REF,
    ALT    = VCF_ALT,
    QUAL   = ".",
    FILTER = ".",
    INFO   = "."
  ) %>%
  distinct() %>%
  arrange(CHROM, POS)

writeLines(c(vcf_lines,
             apply(vcf_data, 1, paste, collapse = "\t")),
           "step03_raw.vcf")

cat("VCF rows written:", nrow(vcf_data), "\n")
```

---

# Step 5 — Normalize VCF (Bash)

The raw VCF is normalized using `bcftools norm` (left-align and split multi-allelic sites),
then compressed with `bgzip` and indexed with `tabix`. Normalization is especially
important because resources such as gnomAD store variants in normalized form, and exact
matching becomes much more reliable after left-alignment and allele standardization.

```{bash step05-normalize, engine="bash"}
bcftools norm \
  -f reference/reference.fa \
  -m -any \
  step03_raw.vcf \
  -o step03_normalized.vcf

bgzip -f step03_normalized.vcf
tabix -p vcf step03_normalized.vcf.gz

echo "Normalization complete."
bcftools stats step03_normalized.vcf.gz | grep "^SN"
```

---

# Step 6 — Build canonical_id Bridge Table (R)

A bridge table is created that maps each original variant (Chr_Start_Ref_Alt in hg19)
to its normalized hg38 representation (`canonical_id`).
The `canonical_id` format is: `chrN-POS-REF-ALT` (hg38, normalized). This is the core
identifier used throughout the rest of the pipeline, because it provides a stable bridge
between the original spreadsheet representation and the normalized representation used for
annotation.

```{r step06-bridge}
df_vcf <- read.csv("step03_pre_bridge.csv")

# Read normalized VCF
norm_vcf <- read.table("step03_normalized.vcf.gz",
                    sep = "\t", comment.char = "#",
                    col.names = c("CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO"))

norm_vcf <- norm_vcf %>%
  mutate(canonical_id = paste(CHROM, POS, REF, ALT, sep = "-"))

# Build bridge: original key → canonical_id
bridge <- df_vcf %>%
  mutate(
    CHROM = paste0("chr", Chr38),
    POS   = VCF_POS,
    REF   = VCF_REF,
    ALT   = VCF_ALT
  ) %>%
  left_join(norm_vcf %>% select(CHROM, POS, REF, ALT, canonical_id),
            by = c("CHROM", "POS", "REF", "ALT")) %>%
  select(canonical_key, canonical_id) %>%
  distinct()

write.csv(bridge, "step05_bridge.csv", row.names = FALSE)
cat("Bridge rows:", nrow(bridge), "\n")
cat("canonical_id NAs:", sum(is.na(bridge$canonical_id)), "\n")
```

---

# Step 7 — Add canonical_id to Main Table (R)

The `canonical_id` from the bridge table is joined to the full variant table.
This enables exact matching against gnomAD records in later steps.

```{r step07-add-canonical}
df <- read_excel("step02_hg38_coords.xlsx", sheet = "variants_hg38")
bridge <- read.csv("step05_bridge.csv")

df <- df %>%
  left_join(bridge, by = "canonical_key")

wb <- createWorkbook()
addWorksheet(wb, "variants_annotated")
writeData(wb, "variants_annotated", df)
saveWorkbook(wb, "step06_with_canonical_id.xlsx", overwrite = TRUE)

cat("Total rows:", nrow(df), "\n")
cat("Rows with canonical_id:", sum(!is.na(df$canonical_id)), "\n")
```

---

# Step 8 — Extract Unique Variants (R)

A deduplicated list of unique variants (by `canonical_id`) is extracted.
This list is used to generate per-chromosome region files for `bcftools query`.

```{r step08-unique}
df <- read_excel("step06_with_canonical_id.xlsx", sheet = "variants_annotated")

unique_variants <- df %>%
  filter(!is.na(canonical_id)) %>%
  select(canonical_id, Chr38, Start38, Ref, Alt, VCF_POS, VCF_REF, VCF_ALT) %>%
  distinct(canonical_id, .keep_all = TRUE)

write.csv(unique_variants, "step07_unique_variants.csv", row.names = FALSE)
cat("Unique variants:", nrow(unique_variants), "\n")
```

---

# Step 9 — Generate Per-Chromosome Region Files (R)

Region files (one per chromosome) are created for efficient `bcftools query`.
Each file contains `CHROM\tPOS\tPOS` lines for the variants on that chromosome.

```{r step09-regions}
unique_variants <- read.csv("step07_unique_variants.csv")

dir.create("query_regions", showWarnings = FALSE)

unique_variants %>%
  mutate(CHROM = paste0("chr", Chr38)) %>%
  group_by(CHROM) %>%
  group_walk(~ {
    region_df <- .x %>%
      transmute(CHROM = CHROM, START = VCF_POS, END = VCF_POS) %>%
      distinct()
    fname <- file.path("query_regions",
                    paste0(unique(.x$CHROM), "_regions.tsv"))
    write.table(region_df, fname,
                sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
  })

cat("Region files created:", length(list.files("query_regions")), "\n")
```

---

# Step 10 — Query gnomAD VCF Files (Bash)

For each chromosome, `bcftools query` extracts allele frequency and homozygosity
data from the gnomAD v3.1.2 genome VCF files.

**Fields extracted:** `CHROM`, `POS`, `REF`, `ALT`, `AF`, `AC`, `AN`, `nhomalt`,
`AF_nfe`, `AC_nfe`, `AN_nfe`, `nhomalt_nfe`

```{bash step10-gnomad-query, engine="bash"}
mkdir -p gnomad_hits

GNOMAD_DIR="gnomad_vcf"
REGIONS_DIR="query_regions"

for CHR in $(seq 1 22) X Y; do
  REGION_FILE="${REGIONS_DIR}/chr${CHR}_regions.tsv"
  VCF="${GNOMAD_DIR}/gnomad.genomes.v3.1.2.sites.chr${CHR}.vcf.bgz"
  OUT="gnomad_hits/chr${CHR}_hits.tsv"

  if [ -f "$REGION_FILE" ] && [ -f "$VCF" ]; then
    bcftools query \
      -R "$REGION_FILE" \
      -f "%CHROM\t%POS\t%REF\t%ALT\t%INFO/AF\t%INFO/AC\t%INFO/AN\t%INFO/nhomalt\t%INFO/AF_nfe\t%INFO/AC_nfe\t%INFO/AN_nfe\t%INFO/nhomalt_nfe\n" \
      "$VCF" > "$OUT"
    echo "chr${CHR}: $(wc -l < $OUT) hits"
  else
    echo "chr${CHR}: skipped (missing file)"
  fi
done
```

---

# Step 11 — Combine gnomAD Hit Files (R)

All per-chromosome gnomAD hit files are merged into a single annotation table.
A `canonical_id` is constructed from the gnomAD output for exact matching.

```{r step11-combine}
gnomad_files <- list.files("gnomad_hits", pattern = "_hits\\.tsv$", full.names = TRUE)

gnomad_cols <- c("CHROM","POS","REF","ALT",
                 "AF","AC","AN","nhomalt",
                 "AF_nfe","AC_nfe","AN_nfe","nhomalt_nfe")

gnomad_all <- map_dfr(gnomad_files, function(f) {
  tryCatch(
    read.table(f, sep = "\t", header = FALSE,
               col.names = gnomad_cols, fill = TRUE,
               quote = "", comment.char = ""),
    error = function(e) NULL
  )
}) %>%
  mutate(
    canonical_id = paste(CHROM, POS, REF, ALT, sep = "-"),
    across(c(AF, AC, AN, nhomalt, AF_nfe, AC_nfe, AN_nfe, nhomalt_nfe),
           ~ suppressWarnings(as.numeric(.)))
  ) %>%
  distinct(canonical_id, .keep_all = TRUE)

write.csv(gnomad_all, "step08_gnomad_annotations.csv", row.names = FALSE)

cat("Total gnomAD variants found:", nrow(gnomad_all), "\n")
cat("Unique canonical_ids:", n_distinct(gnomad_all$canonical_id), "\n")
```

---

# Step 12 — Annotate Main Table with gnomAD Frequencies (R)

The gnomAD annotations are joined to the full variant table using `canonical_id`.
Variants without a gnomAD match (`AF = NA`) represent ultra-rare or novel variants
not observed in the gnomAD v3.1.2 cohort (~750,000 genomes). However, lack of a gnomAD
match should be interpreted as a rarity signal rather than direct evidence of pathogenicity,
so this step supports prioritization but does not replace downstream biological review.

```{r step12-annotate}
df       <- read_excel("step06_with_canonical_id.xlsx", sheet = "variants_annotated")
gnomad   <- read.csv("step08_gnomad_annotations.csv")

gnomad_slim <- gnomad %>%
  select(canonical_id, AF, AC, AN, nhomalt,
         AF_nfe, AC_nfe, AN_nfe, nhomalt_nfe)

df_annotated <- df %>%
  left_join(gnomad_slim, by = "canonical_id")

# QC summary
cat("Total rows:", nrow(df_annotated), "\n")
cat("gnomAD matched rows:", sum(!is.na(df_annotated$AF)), "\n")
cat("gnomAD unmatched (ultra-rare / novel):", sum(is.na(df_annotated$AF)), "\n")
cat("Final column count:", ncol(df_annotated), "\n")

wb <- createWorkbook()
addWorksheet(wb, "variants_gnomad")
writeData(wb, "variants_gnomad", df_annotated)
saveWorkbook(wb, "step09_annotated.xlsx", overwrite = TRUE)
```

---

# Step 13 — Add variant_num and Prepare for Filtration (R)

A new column `variant_num` is added, representing the number of **distinct patients**
carrying each variant (identified by `canonical_id`). This recurrence metric is useful
for prioritizing variants observed in multiple affected individuals.

The `Patient` column is relocated before `canonical_id`, and `variant_num` is placed
immediately after `Patient`. The result is saved as a new sheet `ready_for_filtration`
in the same Excel file, preserving the original `variants_gnomad` sheet. This final
layout is intended to make the next filtering stage more practical by placing patient-level
recurrence information directly next to the main variant identifier.

```{r step13-variant-num}
df <- read_excel("step09_annotated.xlsx", sheet = "variants_gnomad")

# Calculate variant_num: number of distinct patients per canonical_id
variant_num_table <- df %>%
  filter(!is.na(canonical_id), !is.na(Patient)) %>%
  group_by(canonical_id) %>%
  summarise(variant_num = n_distinct(Patient), .groups = "drop")

# Join variant_num back to main table
df <- df %>%
  left_join(variant_num_table, by = "canonical_id")

# Relocate columns: Patient → variant_num → canonical_id → rest
df <- df %>%
  relocate(Patient, .before = canonical_id) %>%
  relocate(variant_num, .after = Patient)

# QC summary
cat("Total rows:", nrow(df), "\n")
cat("Variants seen in >= 2 patients:", sum(df$variant_num >= 2, na.rm = TRUE), "\n")
cat("Variants seen in >= 3 patients:", sum(df$variant_num >= 3, na.rm = TRUE), "\n")
cat("Variants with NA variant_num:", sum(is.na(df$variant_num)), "\n")

# Save as new sheet (keep variants_gnomad intact)
wb <- loadWorkbook("step09_annotated.xlsx")
if ("ready_for_filtration" %in% names(wb)) removeWorksheet(wb, "ready_for_filtration")
addWorksheet(wb, "ready_for_filtration")
writeData(wb, "ready_for_filtration", df)
saveWorkbook(wb, "step09_annotated.xlsx", overwrite = TRUE)

cat("Saved: ready_for_filtration sheet → step09_annotated.xlsx\n")
```

---

# Pipeline File Flow

```
input_variants.xlsx  [Affected only all]
        │
        ▼ Step 1 (R)
liftover_files/step01_bed_hg19.txt
        │
        ▼ Step 2 (Bash: liftOver)
liftover_files/step02_mapped_hg38.txt
        │
        ▼ Step 3 (R)
step02_hg38_coords.xlsx  [variants_hg38]
        │
        ▼ Step 4 (R: Biostrings)
step03_pre_bridge.csv + step03_raw.vcf
        │
        ▼ Step 5 (Bash: bcftools norm + bgzip + tabix)
step04_normalized.vcf.gz
        │
        ▼ Step 6 (R)
step05_bridge.csv  [canonical_id]
        │
        ▼ Step 7 (R)
step06_with_canonical_id.xlsx  [variants_annotated]
        │
        ▼ Step 8 (R)
step07_unique_variants.csv
        │
        ▼ Step 9 (R) → query_regions/chrN_regions.tsv
        │
        ▼ Step 10 (Bash: bcftools query)
gnomad_hits/chrN_hits.tsv
        │
        ▼ Step 11 (R)
step08_gnomad_annotations.csv
        │
        ▼ Step 12 (R)
step09_annotated.xlsx  [variants_gnomad]
        │
        ▼ Step 13 (R)
step09_annotated.xlsx  [ready_for_filtration]  ← final output
```

---

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
